---
title: "A Brief Introduction to Spatial Analysis in R"
date: "04/02/2021"
output:
  html_document: 
    toc: true
    toc_float: true
    collapsed: false
    smooth_scroll: false
---
# Background
The programming language R has been in active development since 1995. While initially R was created for statistical computing, it's popularity, flexibility, and open source development structure has allowed it to grow into something much more diverse. In this tutorial we will showcase using R to import, manipulate, and visualize spatial data.
<br>

## Objective
- introduce spatial libraries (raster, rgdal, sp, tmap)
- read in raster data
- create spatial datasets from a csv
- perform basic spatial analysis
- produce interactive maps
<br>

## How To
This tutorial is meant to be self-paced. To follow along you will need to download the example material from the [Github](https://github.com/GeospatialCentroid/Research-and-Program-Coordinator).

This folder contains a protected areas raster, based on the [Protected Planet](https://www.protectedplanet.net/) world database of protected areas, and a csv of occurrence data of [wild squash data](https://nph.onlinelibrary.wiley.com/doi/full/10.1002/ppp3.10085).

Unzip the folder and store it in a convenient location for you.
<br>

The data being used in this example comes from a 2019 publication that looks at the conservation status of the wild relatives of squash. We will be using it to determine the number of known occurrences that have been observed within a protected area. This is a required step in determining the conservation status of these species. If your interested in more spatial work in R, the entire workflow for the paper can be found [here](https://github.com/dcarver1/cwrSDM)


This material was developed using **R version 3.6.3**. Errors may arise if your version is different.
Check your version by typing r`version` into r console.
If you are interesting in changing your version of R there are a few options describing how [here](https://stackoverflow.com/questions/13656699/update-r-using-rstudio). Also, RStudio allows you to [maintain multiple versions of R](http://derekogle.com/IFAR/supplements/installations/InstallRStudioWin.html#configuring-rstudio).

### Geospatial Data Abstraction Library
[GDAL](https://gdal.org/index.html) is the base library for nearly all spatial data analysis in any computer language. Esri (ArcGIS) and QGIS functions are based on GDAL too. R spatial libraries are based on it, as well as python. As such, it's very important to know this library, but you almost never engage with GDAL directly.

In R, we access GDAL through the **rgdal** library, which we will install now.
```{r eval=FALSE, warning = FALSE}
### We're suggesting the dependencies = TRUE parameter, as we're unsure all functions
### will work properly. This is not a common practice as it will often install
### material that you do not need.
# install.packages("rgdal", dependencies = TRUE)
library(rgdal)
```
<br>
We can view a summary of the package using the following command.
<br>
```{r eval=FALSE, warning = FALSE}

# Type this to view a list of the functions within the library in RStudio
help(package = "rgdal")
```
<br>

If we look at the functions within the library, things appear a little cryptic.
The most readily used function is probably **writeOGR()** which allows you to write out spatial data.


Because GDAL is so foundational, we don't often engage with the functions directly. We rely on other spatial packages which make that connection for us. **raster** for manipulating grid based datasets.  **sp** or **sf** for vector based processing. And there are numerous others for more specific applications.
<br>

>**to SP or SF, that is the question** The honest answer is both! "sp" was established in 2005 and is still the backbone for many other spatial packages in R. "sf" is a newer library and is expected to replace "sp" over time. Maybe the best way to describe the difference is that "sp" creates a spatial object with a dataframe attached. "sf" creates a dataframe with spatial object attached. This means that "sf" objects can be manipulated using the **tidyverse**, which means they behave a bit more like we expect objects in R to behave.
That said, there are times when you need a "sp" object or a "sf" object to enable a specific method to work. I find I lean on "sp", because it's the first one I worked with. Luckly, converting between the object types is easy with functions like **st_to_sf**.

<br>

### Install a Few Spatial Packages
If you've never installed these packages before, you need to uncomment them and run the code. As a general rule, it's best to leave the install.packages() function in your code commented out. This way you don't reinstall the package everytime you run the code.  
<br>

```{r eval = TRUE, warning = FALSE}
#install.packages("raster", dependencies = TRUE)
#install.packages("sp", dependencies = TRUE)
library(raster)
library(sp)
```

<br>

### Loading a Raster into R

We will rely on the raster library to bring in our first bit of spatial data.
<br>

Check to make sure that you are in the correct working directory, which defines where your files are stored/written. It should look something like `~/Desktop/R_SC_Spatial/`. 

Since the lesson is written in an R Project, your working directory should automatically be set to the correct place. R Projects are great, because you can write your script using relative paths and know that the script will work on anyone's computer, because your files are contained within the project. It's like the equivalent of using `setwd()` without having to run `setwd()`. (Read more about that below.)

>**what about setwd()?**  When you set the working directly; you are declaring a relative path from where R studio will look for all files which is  helpful because you can then point to files using r`file <- read.csv("~/example.csv")`. This is very efficient if all your data and outputs are stored within the folder you set as your working directory. So if you want to use setwd() please go ahead.

<br>

```{r eval = TRUE, warning = FALSE}
# read in the file
proLands <- raster::raster(x = "beginnerGeospatialR/proAreas.tif")

### Example with setwd()
# setwd("F:/temp") # this will be unique to your computer
# proLands <- raster::raster(x = ~"/proAreas.tif")

# print it out to view some of the metadata
print(proLands)
```
<br>
The basic plot function can be helpful but it doesn't tell us a ton about the data.

We can view much of the metadata associated with the file just by printing it. Content such as **extent** or **crs** can be selected using functions from the raster library or directly using a special type of indexing that is present in more complex data types (S4 objects) in R.
<br>
```{r eval = TRUE, warning = FALSE}
# use indexing to retrieve specific information
raster::extent(proLands)

# both provide the same information
proLands@extent
```
<br>

We can also use base R to visualize the data.
<br>

```{r eval = TRUE, warning = FALSE}
# quickly visualize the content
plot(proLands)
```
<br>
While this is a spatial file, R is plotting it as it were numeric data. It is simply using the latitude and longitude ranges as a series of coordinates, but spatial coordinates!
<br>




### Generate Spatial Data from a CSV
<br>

```{r eval = TRUE, warning = FALSE}

# read in the data
d1 <- read.csv("beginnerGeospatialR/cucurbitaData.csv")
str(d1)
```
<br>
There are a lot of observations here. Also, we can see there are 3 unique values in the taxon column. We will use those values to subset the data.
<br>

```{r eval = TRUE, warning = FALSE}
# view the unique species
unSpec <- unique(d1$taxon)

### subset all records associated with a species
# select our species of interest by indexing the list
species1 <- unSpec[1]
# filter the occurrence data set for all records where the taxon matches our
# species of interest.
d2 <- d1[d1$taxon == species1,]
 # d1$taxon == species1 returns a vector of TRUE/FALSE values we use to select
 # rows from the d1 dataframe.

 # print a summary of the new subset  
str(d2)
```
<br>

With the dataframe cleaned-up, we will use the **sp** library to generate a spatial object.
<br>

```{r eval = TRUE, warning = FALSE}
### Generate a Spatial Point Dataframe
names(d2)
# coords = df of longitude, latitude (x,y) values
coordinates <- d2[,c(4,3)] # select all rows in the 3rd and 4th columns
# the order is important, (x,y) which equates to (longitude, latitude).
# data = information associated with the records
# proj4string =
pro4 <- proLands@crs # pull the coordinate reference system from the raster layer

#compile all elements into the function.
sp1 <- sp::SpatialPointsDataFrame(coords = coordinates,
                                  data = d2,
                         proj4string = pro4)
# view the object
sp1
```
<br>

We used the **proLands** layer to pull a CRS or coordinate reference system. By doing this, we are assuming that the data from the csv was collected in WGS1984. This is a big assumption, so be aware of it!
<br>

```{r eval = TRUE, warning = FALSE}
# view the data
head(sp1@data)
```
<br>

SpatialPointsDataFrame object has all the original data stored with the spatial elements
<br>

```{r eval = TRUE, warning = FALSE}
plot(sp1)
```
<br>

The plot of the spatial point object does not tell us much. Let's load in a library specifically created for mapping spatial objects to better visual the points.

<br>

```{r eval = TRUE, warning = FALSE}
#install.packages("tmap", dependencies = TRUE)
library(tmap)
# use the quick map function to visualize the data
tmap::qtm(sp1)
```
<br>

Ok, so that didn't change much. We will come back to this and show how to add some complexity and interactivity to our maps with **tmap**.

<br>

### Comparing the Data

<br>

Before we set into an analysis using the two datasets, we will evaluate how well they align
<br>

```{r eval = TRUE, warning = FALSE}
# check extent
raster::extent(proLands) == raster::extent(sp1)
raster::extent(proLands) > raster::extent(sp1)
```
<br>

The extents of the spatial objects are different, and the proLands layer is larger then the point object.
<br>

```{r eval = TRUE, warning = FALSE}
# check coordinate reference system
raster::compareCRS(x = proLands, y = sp1)
```

<br>

As the extents are different, it's worth while to clip the bigger dataset to improve processing time and keep a clean study area.  
<br>

### Crop the Data

<br>
```{r  warning = FALSE}
# Crop the protected areas to the extent of the points
p1Crop <- raster::crop(x = proLands, y = sp1)

#quick visual check with qtm
tmap::qtm(proLands)
```
<br>

```{r warning = FALSE}
tmap::qtm(p1Crop)
```
<br>


The extent has shrunk quite a bit. Now that we know the datasets match spatially, we can conduct an analysis between them.

<br>`

### Extracting Values to Points

<br>

We will use our points and protected lands raster to determine which (if any) of the occurrences are found within protected lands.
<br>

```{r eval = TRUE, warning = FALSE}
# extract values
?extract
```
<br>

```{r warning = FALSE}
# extract returns a vector of length(y), therefore we can just add that
# data as new column to our spatial points dataframe
sp1$inProArea <- raster::extract(x = p1Crop, y = sp1)

head(sp1@data)
```
<br>

With the result recorded in the point data we can visualize the values on a map.
<br>

### Map the Data
<br>

We will start by visualizing the occurrence data on an interactive map using the library tmap.
<br>

```{r warning = FALSE}
# map the points
map <- tmap::tm_shape(shp = sp1)+
  tmap::tm_dots(col = "inProArea", size = 0.1,title = "Occurrences in Protected Areas")
map
```
<br>

The "map" is stored as a tmap object. This is helpful because we can easily add new layers to the existing object.
<br>

```{r warning = FALSE}
# add the raster to the map
map2 <- map +
  tmap::tm_shape(p1Crop) +
    tmap::tm_raster(alpha = 0.4,palette = "green", title = "Protected Areas")
map2
```
<br>
Because we added the protected areas layer on top of the existing map object, we used the visualization parameter 'alpha' to adjust the transparency so we can still see the occurrence data beneath it.

To present this information as a interactive map, we change a 'tmap' setting. We also set the background map. Lastly we did not save this product as a object, so when the code is run, it will execute directly to the terminal.
<br>

```{r warning = FALSE}
# set the map to interactive and adjust the base map
tmap::tmap_mode("view")
map2 + tmap::tm_basemap("OpenStreetMap")
```
<br>
Notice we did not need to change the object type to change how it is displayed. It is the same 'tmap' object from before, now it's just interactive, rather then static.
<br>


While visualizing this information in RStudio is great, it is also helpful to have it as a file, so we can save this map as an image. An important thing to note is that the resulting file is not a spatial layer but a png file.
<br>

```{r eval = FALSE, warning = FALSE}
tmap::tmap_save(tm = map2, filename = "beginnerGeospatialR/map.png")
```
<br>

## Challenge
<br>
Use a for loop and a function to generate the protected points map for each species in the species list.
<br>

```{r eval = FALSE, warning = FALSE}
#hint
for(i in speciesList){
  map <- function(i){
    # subset data
    # read in raster
    # crop raster
    # extract values
    # map content
    return(map)
  }
  tmap::tmap_save(tm = map, filename = paste0(i,"_protectedOccurrences.png"))
}
```
