---
title: "A Brief Introduction to Spatial Analysis in R"
author: "carverd@colostate.edu", "caroline.conley@colostate.edu"
date: "6/15/2020"
output:
  html_document: default
---
# Background
The programming language R has been in active development since 1995. While initially R was created for statistical computing, it's popularity, flexibility, and open source development structure has allowed it to grow into something much more diverse. In this content we will showcase using R to import manipulate and visualize spatial data.
<br>

## Objective
- introduce spatial libraries (raster,rgdal, sp, tmap)
- read in raster data
- create spatial datasets from a csv
- perform basic spatial analysis
- produce interactive maps
<br>

## How to
This content is meant to function as a self based tutorial. To follow along you will need to download the example material form the [Github](https://github.com/dcarver1/R_SC_Spatial/raw/master/exampleData.7z)

This file contains a protected areas raster based on the [Protected Planet](https://www.protectedplanet.net/) world database of protected areas and a csv of occurrence data of [wild squash data](https://nph.onlinelibrary.wiley.com/doi/full/10.1002/ppp3.10085)

Unzip the file and store it in a convenient location for you.
<br>

## at this point it would be helpful to maybe explain what the data is and what the goal of this mapping exercise is "for this exercise, we are going to use this data to map X"

### Geospatial Data Abstraction Library
[GDAL](https://gdal.org/index.html) is the base library for nearly all spatial data analysis in any computer language. ERSI functions are based on it, QGIS functions are based on it. R spatial libraries are based on it. It's very important to know this, but you almost never engage with GDAL directly.

In R, we access GDAL through the **rgdal** library, which we will install now.
```{r eval=FALSE, warning = FALSE}
#install.packages("rgdal")
#This options command was because I kept getting excessive warnings
options("rgdal_show_exportToProj4_warnings"="none")

## I'm going to leave this out for now. I might add it later. I think it's great information but probably a little too deep for the reader. We shall see.
library(rgdal)
```
<br>
We can view a summary of the package using the following command.
<br>
```{r eval=FALSE, warning = FALSE}

#type this to view a list of the functions within the library in RStudio
help(package = "rgdal")
```
<br>

If we look at the functions within the library, things appear a little cryptic.
The most readily used function is probably **writeOGR()** which allows you to write out spatial data.


Because GDAL is so foundational, we don't often engage with the functions directly. We rely on other spatial packages which make that connection for us. **raster** for manipulating grid based datasets.  **sp** or **sf** for vector based processing. There are numerous others for more specific applications.
<br>

>**to SP or SF, that is the question** The honest answer is both. "sp" was established in 2005 and is still the backbone for many other spatial packages in R. "sf" is a newer library and is expect to replace sp over time. Maybe the best way to describe the difference is that "sp" creates a spatial object with a dataframe attached. "sf" creates a dataframe with spatial object attached. This means that "sf" objects can be manipulated using the **tidyverse**, which means they behave a bit more like we expect objects in R to behave.
That said, there are time when you need a "sp" object or a "sf" object to enable a specific method to work. I find I lean on "sp", because it's the first one I worked with. Luckly converting between the object types is easy with functions like **st_to_sf**.

<br>

### Install a Few Spatial Packages
If you've never installed these packages before, you need to uncomment them and run the code. As a general rule it's best to leave the install.packages() function in your code comment out. This way you don't reinstall the package everytime you run the code.  
<br>

```{r eval = TRUE, warning = FALSE}
#install.packages("raster")
#install.packages("sp")
library(raster)
library(sp)
```

<br>

### Loading a raster into R

We will rely on the raster library to bring in our first bit of spatial data.
<br>

```{r eval = TRUE, warning = FALSE}
# set base directory, which defines where your files are store/written
# I am in the habit of setting the working directory, not sure if there is a difference?
setwd("~/Documents/TESTING")

This is a good point. I'll explain the difference
```
<br>

The baseDir show point to where you extract the downloaded data.
<br>

```{r eval = TRUE, warning = FALSE}
# read in the file
proLands <- raster::raster(x = paste0("~/Documents/TESTING/proAreas.tif"))
# print it out to view some of the metadata
# At this point, I got an error saying "Discarded datum Unknown_based_on_WGS84_ellipsoid in CRS definition: +proj=longlat +ellps=WGS84 +towgs84=0,0,0,0,0,0,0 +no_defs,but +towgs84= values preservedDiscarded datum Unknown based on WGS84 ellipsoid in CRS definition,but +towgs84= values preserved" However, it printed as normal and everything else proceeded as it ought to of. not sure what that is but figured i would report it anyways.
print(proLands)
```
<br>
The basic plot function can be helpful but it doesn't tell us a ton about the data.

We can view much of the meta data associated with the file just by printing it. Content such as **extent** or **crs** can be selected using functions from the raster library or directly using a special type of indexing that is present in more complex data types(S4 objects) in R.
<br>
```{r eval = TRUE, warning = FALSE}
#use indexing to retreive specific information
raster::extent(proLands)

# both provide the same information
proLands@extent
```
<br>

We can also use base R to visualize the data.
<br>

```{r eval = TRUE, warning = FALSE}
#quickly visualize the content
plot(proLands)
```
<br>
While this is a spatial file, R is plotting it as it word numeric data. It is simply using the latitude and longitude ranges as a series of coordinates, but spatial coordinates
<br>




### Generate Spatial Data from a csv
<br>

```{r eval = TRUE, warning = FALSE}

# read in the data
d1 <- read.csv(paste0("~/Documents/TESTING/cucurbitaData.csv"))
str(d1)
```
<br>
There are a lot of observation here. Also we can see there area 3 unique values in the taxon column. We will use those values to subset the data.
<br>

```{r eval = TRUE, warning = FALSE}
#view the unique species
unSpec <- unique(d1$taxon)

# subset all records associated with a species
d2 <- d1[d1$taxon == unSpec[1],]
str(d2)
```
<br>

With the dataframe clean up we will use the **sp** library to generate a spatial object.

## this part was a bit over my head, but maybe I am not the norm
Ok good to note. I'll break out the indexing and introduce some more complete comments. The goal will be one action per line of code.
<br>

```{r eval = TRUE, warning = FALSE}
# generate a spatial point dataframe
# coords = df of longitude , latitude (x,y) values
# data = information associate with the records
# proj4string =
sp1 <- sp::SpatialPointsDataFrame(coords = d2[,c(4,3)],
                                  data = d2,
                         proj4string = proLands@crs)
# view the object
sp1
```
<br>

We used the **proLands** layer to pull a CRS or coordinate reference system. By doing this we are assuming that the data from the csv was collected in WGS1984.
<br>

```{r eval = TRUE, warning = FALSE}
# view the data
head(sp1@data)
```
<br>

SpatialPointsDataFrame object has all the original data store with the spatial elements
<br>

```{r eval = TRUE, warning = FALSE}
plot(sp1)
```
<br>

The plot of the spatial point object does not tell us much. Lets load in a library specifically created for mapping spatial objects to better visual the points

<br>

```{r eval = TRUE, warning = FALSE}
#install.packages("tmap")
#install.packages("XML")
# you need XML to run tmap and I previously did not have this package
# also, I was generally having trouble with some of these packages and then I updated my R to the newest version and that helped (so you may want to add in something about needing a more updated version of R)
library(xml2)
Ok, good note about the R version. I'll add a writen warning at the top saying that this was develop in version 3.63. Error may arise if you are using a different version.

For the libraries, we will add the install dependences == true command and that should help with the package issues.
library(tmap)
# use the qick map function to visualize the data
tmap::qtm(sp1)
```
<br>

Ok, so that didn't change much. We will come back to this and so how to add some complexity and interactivity to our maps with **tmap**.

<br>

### Comparing the data

<br>

Before we set into an analysis using the two datasets we will evaluate how well they align
<br>

```{r eval = TRUE, warning = FALSE}
# check extent
extent(proLands) == extent(sp1)
extent(proLands) > extent(sp1)
```
<br>

The extents of the spatial objects are different and the proLands layer is large then the point object.
<br>

```{r eval = TRUE, warning = FALSE}
# check coordinate reference system
raster::compareCRS(x = proLands, y = sp1)
```

<br>

As the extents are different it's worth while to clip the bigger dataset to improve processing time and keep a clean study area.  
<br>

### Crop the data

<br>
```{r  warning = FALSE}
# Crop the protected areas to the extent of the points
p1Crop <- raster::crop(x = proLands, y = sp1)

#quick visual check with qtm
# it seemed like qtm wasn't compatible with the newest version of R so I had to get the newer version
install.packages("qtm2")
library(qtm)
interesting. I'm not familiar with qtm2? I'll include reference to all function librarys. so qtm() will become tmap::qtm() this should help
qtm(proLands)
```
<br>

```{r warning = FALSE}
qtm(p1Crop)
```
<br>


The extent has shrunk quite a bit. Now that know the datasets match spatially we can conduct an analysis between them.

<br>`

### Extracting values to points

<br>

We will use our points and protect lands raster to determine which if any of the occurrence are found within protect lands.
<br>

```{r eval = TRUE, warning = FALSE}
# extract values
?extract
```
<br>

```{r warning = FALSE}
# extract returns a vector of length(y), therefore we can just at that
# data as new column to our spatial points dataframe
sp1$inProArea <- raster::extract(x = p1Crop, y = sp1)

head(sp1@data)
```
<br>

With the result recorded in the point data we can visualize the values on a map.
<br>

### Map the data
<br>

We will start by visualizing the occurrence data on an interactive map using the library tmap.
<br>

```{r warning = FALSE}
# map the points
map <- tm_shape(shp = sp1)+
  tm_dots(col = "inProArea", size = 0.1,title = "Occurrences in Protected Areas")
map
```
<br>

The "map" is store as a tmap object. This is helpful because we can easily add new layers to the existing object.
<br>

```{r warning = FALSE}
# add the raster to the map
map2 <- map +
  tm_shape(p1Crop) +
    tm_raster(alpha = 0.4,palette = "green", title = "Protected Areas")
map2
```
<br>
Because we added the protected areas layer on top of the existing map object, we used the visualization parameter alpha to adjust the transparency so we can still see the occurrence data beneath it.

To present this information as a interactive map, we change a tmap setting. We also set the background map. Lastly we did not save this product as a object, so when the code is ran it will execute directly to the terminal.
<br>

```{r warning = FALSE}
# set the map to interactive and adjust the base map
# you need the package knitr to do this which I didn't have previously
#install.packages("knitr")
library(knitr)
Perfect thank you
tmap::tmap_mode("view")
map2 + tm_basemap("OpenStreetMap")
```
<br>
Notice we did not need to change the object type to change how it is displayed. It is the same tmap object from before, now it's just an interactive rather then static map.
<br>


While visualizing this information in R studio it great it will be helpful to have it as a file. We can save this map as an image. As important thing to not is that the resulting file is not a spatial layer but a png.
<br>

```{r eval = FALSE, warning = FALSE}
tmap::tmap_save(tm = map2, filename = paste0("~/Documents/TESTING/map.png"))
```
<br>

## challenge
<br>
Use a for loop and a function to generate the protected points map for each species in the species list.
<br>

```{r eval = FALSE, warning = FALSE}
#hint
for(i in speciesList){
  map <- function(i){
    # subset data
    # read in raster
    # crop raster
    # extract values
    # map content
    return(map)
  }
  tmap::tmap_save(tm = map, filename = paste0(i,"_protectedOccurrences.png"))
}
```
